
# ME314 Introduction to Data Science and Machine Learning

## Day 3 Assignment, LSE ME314 2021
---

### 1.  Normalizing data

This question uses this table:

   | EmployeeID | SalesPerson | SalesOffice | OfficeNumber | Customer1 | Customer2 | Customer3 |
   | ------:|:-------------:|:----------:|:------------ |:-------:|:------:|:------:|
   | 1003   | Mary Smith    |  Chicago   | 312-555-1212 | Ford   | GM |       |
   | 1004   | John Hunt     |  New York  | 212-555-1212 | Dell   | HP | Apple |
   | 1005   | Martin Hap    |  Chicago   | 312-555-1212 | Boeing |    |       |

from the lecture notes.  For each answer, please explain as fully as possible, and feel free to use tables or diagrams if you prefer.

a)  Why does this table violate the first normal form, and what would be required to make it 1NF?

First, let's convert this markdown table into a tibble.

```{r, message=FALSE,warning=FALSE}

library('janitor')
library('tidyverse')

rmd.string <-
  '| EmployeeID | SalesPerson | SalesOffice | OfficeNumber | Customer1 | Customer2 | Customer3 |
   | ------:|:-------------:|:----------:|:------------ |:-------:|:------:|:------:|
   | 1003   | Mary Smith    |  Chicago   | 312-555-1212 | Ford   | GM |       |
   | 1004   | John Hunt     |  New York  | 212-555-1212 | Dell   | HP | Apple |
   | 1005   | Martin Hap    |  Chicago   | 312-555-1212 | Boeing |    |       |'

rmd.table <- read_delim(rmd.string, delim = '|')[-1,-c(1,9)] 

colnames(rmd.table) <- str_squish(colnames(rmd.table))

rmd.table <- 
  rmd.table %>% 
  mutate(EmployeeID = as.integer(EmployeeID),
         across(SalesPerson:Customer3, ~ str_squish(.x)))

rmd.table %>% 
  pivot_longer(cols = starts_with('Customer'),
               names_to = NULL,
               values_to = 'Customer') %>% 
  filter(Customer != '')

```

We can see that customer information is contained over three `Customer` columns. To make sure the database satisfies the 1NF, let's create a column for customers by pivoting the data.

```{r}

table.1nf <-
  rmd.table %>% 
  pivot_longer(cols = starts_with('Customer'),
               names_to = NULL,
               values_to = 'Customer') %>% 
  filter(Customer != '')

```

b)  What additional steps would be needed to make the table 2NF? What about 3NF?

Let's break up our table into employees and customers.

```{r}

employees <-
  table.1nf %>% 
  distinct(EmployeeID, SalesPerson, SalesOffice, OfficeNumber)

employees

customers <-
  table.1nf %>% 
  select(EmployeeID, Customer)

customers

```

As all non-key columns depend on the table's primary key, 2NF is satisfied. However, because there is a transitive dependency (i.e. one that goes through another attribute) 3NF is violated. In this case, because we know that `EmployeeID` can be used to find `OfficeNumber` which can, in turn, be used to find `SalesOffice`. To ensure 3NF is not violated, we could further break `employees` down into two further databases.

```{r}

offices <-
  employees %>% 
  select(OfficeNumber, SalesOffice)

offices

employees <-
  employees %>% 
  select(EmployeeID, SalesPerson, OfficeNumber)

employees

```

c)  Why might we not want to normalize data to the fullest extent possible?

If your database is relatively small, it may be more efficient to look at variables in a more integrated format. However, for larger databases, storing information in different tables will be more efficient.

d)  In the table below, which of the three normalization rules does this violate, if any, and why?

   |  countryID  |  countryName    |   EUmember   |  EUjoindate  |
   | -----------:|:----------------|:------------:|:------------:|
   | 00001       | France          |  `true`      |  1958-01-01  |
   | 00004       | Hungary         |  `true`      |  2004-05-01  |
   | 00003       | Serbia          |  `false`     |       `NULL` |
   | 00004       | Finland         |  `true`      |  1995-01-01  |
   | 00005       | Russia          |  `false`     |       `NULL` |
   | 00006       | Ireland, UK     |  `true`      |  1973-01-01  |

   Are there any other problems with the table, besides normalization?

Two rules are violated:

- 1NF is violated because there are two values in the last row of `countryName`.
- 3NF is violated because `EUmember` can be determined by whether `EUjoindate` is not `NULL`.

e)  What would it take to fully (1NF-3NF) normalize this dataset?

   Write out these tables, and describe why this meets each of the normal forms.  This is a database of movies watched on NetBricks, a streaming movie service.

   | Name           | Address    |   Movies Rented   |  Salutation  | Category | Fiction? |
   |:---------------|:-----------|:------------------|:------------:|----------|----------|
   | Bob Smith      | 1 Houghton Street  | _Star Wars_, _Fast and Furious 7_ |  Dr. |  Scifi, Action | Yes, Yes|
   | Pry Ministair  | 10 Downing St  |  _Brexit: Three Years On_   |  Rt Hon.  | Documentary | No |
   | Joe Bloggs     | 8 Myhatt St.  |  _The Big Short_, _Star Wars_  | Mr. | Documentary, Scifi | No, Yes |

First, let's convert this markdown table to a tibble.

```{r, message=FALSE,warning=FALSE}

rmd.string <-
'   | Name           | Address    |   Movies Rented   |  Salutation  | Category | Fiction? |
   |:---------------|:-----------|:------------------|:------------:|----------|----------|
   | Bob Smith      | 1 Houghton Street  | _Star Wars_, _Fast and Furious 7_ |  Dr. |  Scifi, Action | Yes, Yes|
   | Pry Ministair  | 10 Downing St  |  _Brexit: Three Years On_   |  Rt Hon.  | Documentary | No |
   | Joe Bloggs     | 8 Myhatt St.  |  _The Big Short_, _Star Wars_  | Mr. | Documentary, Scifi | No, Yes |'

rmd.table <- read_delim(rmd.string, delim = '|')[-1,-c(1,8)] 

colnames(rmd.table) <- str_squish(colnames(rmd.table))

rmd.table <- 
  rmd.table %>% 
  mutate(across(everything(), ~ str_squish(str_remove_all(.x, '\\_')))) %>% 
  janitor::clean_names()

```

To conform to 1NF, let's make sure each column represents a single attribute.

```{r}

rmd.table <-
  rmd.table %>% 
  separate_longer_delim(c(movies_rented, category, fiction), delim = ', ')

rmd.table$userID <- c('001','001','002','003','003')
rmd.table$movieID <- as.character(c(1,2,3,4,1))

rmd.table

```

Now let's create separate tables for `rentals`, `users`, and `movies` to satisfy 2NF.

```{r}

rentals <-
  rmd.table %>% 
  distinct(userID, movieID)

rentals

users <-
  rmd.table %>% 
  distinct(userID, name, address, salutation)

users

movies <-
  rmd.table %>% 
  distinct(movieID, movies_rented, category, fiction)

movies

```

As there is a transitive dependency between `fiction` and `category`, we can create a separate category table.

```{r}

categories <-
  movies %>% 
  distinct(category, fiction)

categories

movies <-
  movies %>% 
  select(-fiction)

movies

```

### 2.  Reshaping data

For this exercise, we will use the **nycflights13** R package, whose tables have been output in `.csv` form [here](nycflights13/).  You may do the following in either R or Python.  Note that this example is developed extensively in [_R for Data Science_](http://r4ds.had.co.nz/relational-data.html).

a)  Create a subtable of the `flights` data, that departed before 05:53 on 2013-02-28.  How many rows and columns does this subtable have?  

```{r, message=FALSE,warning=FALSE}
library('tidyverse')
library("lubridate")

flights <- read_csv("nycflights13/flights.csv")[,-1]

flights$dep_datetime <- ymd_hm(sprintf("%s-%02d-%02d %02d:%02d",
                            flights$year,
                            flights$month,
                            flights$day,
                            flights$hour,
                            flights$minute))
flights_sub <- subset(flights, dep_datetime < ymd_hm("2013-02-28 05:53"))

nrow(flights_sub)
```

b)  Merge or join the subtable from a. `flights` data, to produce a result that includes:  
   *  Departure time
   *  Carrier (two digit code, from `flights`)
   *  Flight number
   *  Destination airport name (hint: you will need to get this from the `airports` table)  

```{r, message=FALSE,warning=FALSE}

airports <- read_csv("nycflights13/airports.csv")[,-1]

flights %>% 
  left_join(airports, by = c('dest' = 'faa')) %>% 
  select(dep_datetime, carrier, flight, name)
  
```

c) For every airline that had flights in the `flights` data compute the average age of the planes it flew from the entire dataset.  Age here will be defined as 2013 minus the `year` variable from the `planes` data.  Hint: This involves a join operation on `tailnum`, but also a grouped mean to compute the age (and subtracting 2013, which you can do before or after the computation of the mean).


```{r, message=FALSE,warning=FALSE}

planes <- read_csv("nycflights13/planes.csv")[,-1]
airlines <- read_csv("nycflights13/airlines.csv")[,-1]

flights %>% 
  distinct(tailnum, carrier) %>% 
  left_join(planes, by = 'tailnum') %>% 
  group_by(carrier) %>% 
  summarise(mean_age = mean(2013 - year, na.rm = TRUE)) %>% 
  left_join(airlines, by = 'carrier') %>% 
  select(name, mean_age)

```

### 3.  Groupwork
Get into pairs or groups and think of a research question you might have using the flights data and produce some output to discuss (no more than 2-3 minutes) at the start of the next class. (You can send the code via slack to be run by Sarah or Yuanmo or connect your own laptop.)

Consult the dplyr reference of the many functions that there are: 
[dplyr reference](https://dplyr.tidyverse.org/reference/index.html)


```{r}
## your code
```
